---
title: "Where are the cane toads going?"
format: html
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(readr)
library(fs)
lapply(dir_ls("R", glob = "*.R"), source)
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

front <- read_csv("output/toad-front.csv")
summary_stats <- read_csv("output/toad-summary.csv")
```

Cane toads were released near Gordonvale in 1935. This is where they had got to
in `r basename(summary_stats$source)`, using `r format(summary_stats$records_clean, big.mark = ",")`
records from the Atlas of Living Australia.

## How fast is the front moving?

**`r round(summary_stats$km_per_year, 1)` kilometres a year.**

```{r}
#| label: fig-front
#| fig-cap: "The westernmost record reached by the end of each decade."
plot_front(front)
```

## How much data is there?

```{r}
#| label: fig-records
#| fig-cap: "Records per decade. Most of them are recent."
plot_decade(front)
```

Of `r format(summary_stats$records, big.mark = ",")` records supplied,
`r format(summary_stats$records - summary_stats$records_clean, big.mark = ",")`
were dropped as impossible: coordinates outside Australia, or dates before the
toads arrived.
